---
title: "Deutsch-Sozsa Algorithm"
author: "Carsten Urbach"
output:
  rmarkdown::html_vignette
    
vignette: >
  %\VignetteIndexEntry{Deutsch-Sozsa Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE}
library(knitr)
library(qsimulatR)
```

# The Deutsch-Sosza Algorithm

This is an example implementation of the Deutsch-Sosza algorithm for
the special case of $2$ qbits. The algorithm allows to distinguish
between a constant or balanced function $f$ with a single application
of $f$, relying on what is called quantum parallelism.

We first prepare a state $|\psi_0\rangle = |x, y\rangle = |0, 1\rangle$ with only 2
qbits as follows

```{r}
x <- X(1) * qstate(nbits=2, basis=genComputationalBasis(2, collapse=","))
x
```

Note that we count the qbits from one to number of qbits, and the **least
significant bit** (the right most one) is counted **first**. 

Using the Hadamard gate on both qbits results in a superposition in both qbits

```{r}
y <- H(2) * (H(1) * x)
y
```

The next step is to apply the uniform transformation $U_f$ to the
state $|x\rangle(|0\rangle - |1\rangle)$. The action of $U_f$ was
defined as $|x,y\rangle \to |x, y\oplus f(x)\rangle$, where $\oplus$
is addition modulo $2$. The function $f$ is a function
$\{0,1\}\to\{0,1\}$. 

We first consider a so-called balanced function $f(x)$, i.e.\ it is equal to
$1$ for exactly half of the possible $x$. In our case with a single
qbit $x$ this could be $f(0)=0$ and $f(1) = 1$. 

$U_f$ is realised in this case by CNOT$(2,1)$, where we consider the
second qbit as the control qbit. For $|x, y\oplus f(x)\rangle$, there
are four different possiblities

* $x=0, y=0$, $U_f(|0,0\rangle) = |0, 0\oplus f(0)\rangle = |0, 0\rangle$
* $x=1, y=0$, $U_f(|1,0\rangle) = |1, 0\oplus f(1)\rangle = |1, 1\rangle$
* $x=0, y=1$, $U_f(|0,1\rangle) = |0, 1\oplus f(0)\rangle = |0, 1\rangle$
* $x=1, y=1$, $U_f(|1,1\rangle) = |1, 1\oplus f(1)\rangle = |1, 0\rangle$

Now, 

* CNOT$(2,1)|0,0\rangle = |0,0\rangle$
* CNOT$(2,1)|1,0\rangle = |1,1\rangle$
* CNOT$(2,1)|0,1\rangle = |0,1\rangle$
* CNOT$(2,1)|1,1\rangle = |1,0\rangle$

which is what we wanted to archieve. Thus, we apply it:

```{r}
z <- CNOT(c(2, 1)) * y
z
```

Now apply the Hadamard gate again on $x$ (the query register),
i.e. the second qbit 

```{r}
u <- H(2) * z
u
```

Now qbit $2$ equals $1$, thus, if we measure, 

```{r}
value <- measure(u, 2)$value
```

we obtain $`r value`$. We can also plot the corresponding circuit

```{r, fig.align='center'}
plot(measure(u, 2)$psi, qbitnames=c("|y>", "|x>"), cbitnames="c")
```



On the other hand, a constant function
$f(x) = 1$ leads to

* $x=0, y=0$, $U_f(|0,0\rangle) = |0, 0\oplus f(0)\rangle = |0, 1\rangle$
* $x=1, y=0$, $U_f(|1,0\rangle) = |1, 0\oplus f(1)\rangle = |1, 1\rangle$
* $x=0, y=1$, $U_f(|0,1\rangle) = |0, 1\oplus f(0)\rangle = |0, 0\rangle$
* $x=1, y=1$, $U_f(|1,1\rangle) = |1, 1\oplus f(1)\rangle = |1, 0\rangle$

which can be realised with a NOT operation on the first qbit

* X$(1)|0,0\rangle = |0,1\rangle$
* X$(1)|1,0\rangle = |1,1\rangle$
* X$(1)|0,1\rangle = |0,0\rangle$
* X$(1)|1,1\rangle = |1,0\rangle$

So, the same algorithm again, now with the constant $f$

```{r}
x <- X(1) * qstate(nbits=2, basis=genComputationalBasis(2, collapse=","))
y <- H(2) * (H(1) * x)
z <- X(1) * y
z
u <- H(2) * z
u
value <- measure(u, 2)$value
```

and we obtain $`r value`$ for the second qbit.

