---
title: "Exponentiation modulo n"
author: "Carsten Urbach"
output:
  rmarkdown::html_vignette
    
vignette: >
  %\VignetteIndexEntry{Exponentiation modulo n}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE}
library(knitr)
library(qsimulatR)
knitr::opts_chunk$set(fig.align='center',
                      comment='')
```

Note that for this we roughly follow the paper by Vedral, Barenco and
Ekert (arXiv:9511018). However, we use the addition by qft instead of
the procedure using Toffoli gates they propose.

## Multiplication modulo $n$

For multiplying a state $|j\rangle$ with a constant $a$ we can follow
the scheme to multiply two numbers in binary representation. As an
example, multiply $5$ with $3$. $5$ has binary representation $0101$
and $3$ has $0011$. So, the procedure for $5\cdot 3$ is

```
  00011 * 1 (1*3)
+ 00110 * 0 (0*6)
+ 01100 * 1 (1*12)
+ 11000 * 0 (0*24)
= 01111 (15)
```

Now, if we have a controlled add operation, we can use the qubits of
the first register (in this case representing $5$) as control bits and
the other register as the constant to add. The single terms in the sum
can be efficiently pre-computed classically as follows

```{r}
summands <- function(x, n, N) {
  b <- as.integer(intToBits(x))
  ret <- c()
  for(i in c(1:N)) {
    s <- 0
    for(j in c(1:N)) {
      s <- s+as.integer(b[j])*2^(i+j-2)
    }
    ret[i] <- s %% n
  }
  return(ret)
}
```

Now we need a controlled add operation. Here we can build on our add
operation using the qft for which we need a controlled phase shift
operation first

```{r}
cRtheta <- function(bits, theta=0.) {
  cqgate(bits=bits, gate=methods::new("sqgate", bit=as.integer(bits[2]),
                                      M=array(as.complex(c(1, 0, 0, exp(1i*theta))),
                                              dim=c(2,2)), type="Rt"))
}
```

```{r}
cadd <- function(c, bits, x, y) {
  n <- length(bits)
  z <- cqft(c=c, x=x, bits=bits)
  for(i in c(1:n)) {
    z  <- cRtheta(bits=c(c, bits[i]), theta = 2*pi*y/2^(n-i+1)) * z
  }
  z <- cqft(c=c, x=z, inverse=TRUE, bits=bits)
  return(invisible(z))
}
```

Let's check whether it works as expected:

```{r}
basis <- c()
for(i in c(0:(2^4-1))) basis[i+1] <- paste0("|", i %/% 2, ">|", i %% 2, ">")
x <- H(1)*qstate(4, basis=basis)
c <- 1
bits <- c(2:4)
z <- cadd(c=c, bits=bits, x=x, y=5)
z
z <- cadd(c=c, bits=bits, x=z, y=2)
z
```

Equipped with this functionality, we can finaly perform a binary
multiplication. Note that we need two registers, the first one to
store the inital value and the second one to store the final result of
the multiplication

```{r}
mult <- function(reg1, reg2, x, y, swap=TRUE) {
  stopifnot(length(reg1) == length(reg2))
  n <- length(reg2)
  s <- summands(y, 2^n, n)
  for(i in c(1:n)) {
    x <- cadd(c=reg1[i], bits=reg2, x=x, y=s[i])
  }
  if(swap) {
    for(i in c(1:n)) {
      x <- SWAP(c(reg1[i], reg2[i])) * x
    }
  }
  return(invisible(x))
}
```

With this we can perform a _reversible_ multiplication, which is why
we we introduced the SWAP operations at the end. They interchange the
two registers. The result is the following

```{r}
basis <- c()
for(i in c(0:(2^3-1))) {
  for(j in c(0:(2^3-1))) {
    basis[i*2^3+j + 1] <- paste0("|", i, ">|", j, ">")
  }
}
x <- X(2)*qstate(6, basis=basis)
x
reg1 <- c(1:3)
reg2 <- c(4:6)
z <- mult(reg1, reg2, x=x, y=3)
z <- X(5) * z
z
```

Let's be a bit more precise here for the multiplication of, say $a$
and $b$. Starting with both registers in state $|0\rangle$, we can
first bring the result register to state $|1\rangle$ using a NOT
operation, i.e. $|0\rangle|1\rangle$. Now, we multiply by $a$, which
leaves us with state 
\[
|a \mod n \rangle|1\rangle
\]
which we can apply the NOT gate again to reset the result register to
state $|0\rangle$ and then we swap the two registers to arrive at
\[
|0\rangle|a \mod n\rangle\,.
\]
Now we multiply by $b$ getting us to state
\[
|a\times b \mod n\rangle|a \mod n\rangle\,.
\]
Multiply the result register with the inverse of $a \mod n$ and apply
the NOT gate getting us to 
\[
|0\rangle|a\times b \mod n\rangle\,.
\]
The inverse modulo $n$ can be computed efficiently in a classical
way by the extended Euclidean algorithm

```{r}
eEa <- function(a, b) {
  if(a == 0) return(c(b, 0, 1))
  res <- eEa(b %% a, a)
  return(c(res[1], res[3] - (b %/% a) * res[2], res[2]))
}

moduloinverse <- function(a, n) {
  res <- eEa(a=a, b=n)
  if(res[1] != 1) stop("inverse does not exist!")
  return(res[2] %% n)
}
```

If $a$ and $n$ are not coprime, the iverse does not exist. However,
for the application we have in mind this is not an issue.

## Working modulo $N$

So far we have assumed that we work modulo $n$, where $n$ was dictated
by the number of qubits. However, this is not the realistic case. We
have to write an adder modulo $N$, i.e. $|x\rangle\to|x + y\mod
N\rangle$. We can implement this by subtracting $N < n$ whenever
needed. We will follow the convention that if $x\geq N$ the operation
will be $|x\rangle\to|x\rangle$. Moreover, we will assume $x,y\geq 0$.

To find out, when this subtraction is needed, is a bit tricky. We want
to add $y$ to $x$. To decide beforehand, whether or not we have to
subtract $N$, we have to check whether $x < N-y$. If not, we have to
subract $N$. If we subtract $N-y$ from this state $|x\rangle$, the most
significant qubit indicates whether there occured an overflow. Using a
CNOT gate we can store this info in one ancilla bit $c_1$. The we add
$N-y$ again to retain the original state. Such an operation can be
implemented as follows in a controlled manner (control bit $c$, `bits`
the bits in state $|x\rangle$ where x is stored, $a$ an ancilla bit
and $y$ the value to compare with.

```{r}
cis.less <- function(c, bits, x, c1, a, y) {
  ## add ancilla bit as most significant bit to bits
  b <- c(bits, a)
  n <- length(b)
  ## cadd works modulo 2^n
  z <- cadd(c=c, bits=b, x=x, y=2^n-y)
  ## 'copy' overflow bit
  z <- CNOT(c(a, c1)) * z
  ## add back, resetting a to |0>
  z <- cadd(c=c, bits=b, x=z, y=y)
  return(z)
}
```
This routine will set the qubit $|c_1\rangle$ to 1 if
$|x_\mathrm{bits}\rangle$ is smaller than $y$ and leave it at zero
otherwise. It uses $|a\rangle$ as ancilla bit and $|c\rangle$ as
control bit. Here an example

```{r}
basis <- c()
for(i in c(0:(2^6-1))) {
  basis[i + 1] <-
    paste0("|", i %/% 8 ,">|a=",
    (i %/% 4) %% 2, ">|c1=", (i%/%2) %% 2,
    ">|c=", i%%2, ">")
}

x <- X(1)*qstate(6, basis=basis)
z <- cadd(c=1, bits=c(4,5,6), x=x, y=5)
z
## 5 < 7 -> c1 = 1
v <- cis.less(c=1, bits=c(4,5,6), x=z, c1=2, a=3, y=7)
v
## 5 > 3 -> c1 = 0
w <- cis.less(c=1, bits=c(4,5,6), x=z, c1=2, a=3, y=3)
w
```

Now, recall that if $x\geq N$ we want the operation to leave the state
unchanged. So, we need two `cis.less` operations, one to check whether
$x < N$, which we store in ancilla qubit $c_1$ and another one to
check whether $x < N-y$ stored in $c_2$. Note that the combination
$c_1=0, c_2=1$ is not possible. The implementation looks as follows:

```{r}
caddmodN <- function(c, bits, c1, c2, a, x, y, N) {
  stopifnot(length(a) == 1 && length(c1) == 1 &&
            length(c2) == 1 &&
            length(unique(c(c1, c2, a))) == 3)
  ## set c1=1 if x < N
  z <- cis.less(c=c, bits=bits, x=x, c1=c1, a=a, y=N)
  ## set c2=1 if x < N - y
  z <- cis.less(c=c, bits=bits, x=z, c1=c2, a=a, y=N-y)

  ## if c1 and not c2, x = x + y - N
  z <- X(c2) *( CCNOT(c(c1, c2, a)) * (X(c2) * z))
  z <- cadd(c=a, bits=bits, x=z, y=y-N)
  z <- X(c2) * (CCNOT(c(c1, c2, a)) * (X(c2) * z))

  ## if c1 and c2 add x = x + y
  z <- CCNOT(c(c1, c2, a)) * z
  z <- cadd(c=a, bits=bits, x=z, y=y)
  z <- CCNOT(c(c1, c2, a)) * z

  ## reset c1,2
  z <- cis.less(c=c, bits=bits, x=z, c1=c2, a=a, y=y)
  z <- CNOT(c(c1, c2)) * z
  z <- cis.less(c=c, bits=bits, x=z, c1=c1, a=a, y=N)
  return(invisible(z))
}
```

For the reset part: in the first step we flip $c_2$ if $x + y \mod N
< y$. This can only be true if $x<N-y$. If $c_2$ was 1, its zero now
and the other way around. The next CNOT gate flips $c_2$
to $|0\rangle$. The last step resets $c_1$ to $|0\rangle$.

You can also see from the routine above that, if $x\geq N$ then
`caddmodN` leaves the state unchanged.

Example

```{r}
basis <- c()
for(i in c(0:(2^7-1))) {
  basis[i + 1] <-
    paste0("|", i %/% 16 , ">|a=", (i %/% 8) %% 2,
           ">|c1=", (i %/% 4) %% 2,
           ">|c2=", (i%/%2) %% 2, ">|c=", i%%2, ">")
}

x <- X(1)*qstate(7, basis=basis)
x
bits <- c(5,6,7)
c <- 1
c1 <- 2
c2 <- 3
a <- 4
N <- 7
z <- caddmodN(c=c, bits=bits, c1=c1, c2=c2, a=a, x=x, y=3, N=N) # 0 + 3 mod 7
z
z <- caddmodN(c=c, bits=bits, c1=c1, c2=c2, a=a, x=z, y=5, N=N) # 3 + 5 mod 7
z
```

# Controlled Multiplier modulo $N$

Now, like the `mult` function above a version performing
$|x\rangle\to|x+y\mod N\rangle$. Here we also include the
back-computation of the second register. `reg1` is the result
register. 

```{r}
cmultmodN <- function(c, reg1, reg2, ancillas, x, y, N) {
  stopifnot(length(reg1) == length(reg2))
  ## need 4 ancilla registers
  stopifnot(length(ancillas) == 4 &&
            length(unique(ancillas)) == 4)
  n <- length(reg2)
  ## precompute terms in the sum
  s <- summands(y, 2^n, n)
  for(i in c(1:n)) {
    x <- CCNOT(c(c, reg1[i], ancillas[4])) * x
    x <- caddmodN(c=reg1[i], bits=reg2,
                  c1=ancillas[1], c2=ancillas[2],
                  a=ancillas[3],
                  x=x, y=s[i] %% N, N=N)
    x <- CCNOT(c(c, reg1[i], ancillas[4])) * x
  }
  for(i in c(1:n)) {
    x <- CSWAP(c(c, reg1[i], reg2[i])) * x
  }
  yinv <- N - moduloinverse(a=y, n=N)
  s <- summands(yinv, 2^n, n)
  for(i in c(1:n)) {
    x <- CCNOT(c(c, reg1[i], ancillas[4])) * x
    x <- caddmodN(c=reg1[i], bits=reg2,
                  c1=ancillas[1], c2=ancillas[2],
                  a=ancillas[3],
                  x=x, y=s[i] %% N, N=N)
    x <- CCNOT(c(c, reg1[i], ancillas[4])) * x
  }
  return(invisible(x))
}
```

Example with two 3 qubit registers, which is starting to become slow,
because in total we need 11 qubits

```{r}
basis <- c()
for(i in c(0:(2^11-1))) {
  basis[i + 1] <-
    paste0("|reg1=", i %/% (32*2^3) , ">|reg2=", (i %/% 32) %% 2^3 ,
           "|t=", (i %/% 16) %% 2,
           ">|a=", (i %/% 8) %% 2, ">|c1=", (i %/% 4) %% 2,
           ">|c2=", (i%/%2) %% 2, ">|c=", i%%2, ">")
}
x <- X(9) * (X(1)*qstate(11, basis=basis))
x
c <- 1
ancillas <- c(2:5)
reg2 <- c(6:8)
reg1 <- c(9:11)
N <- 5
z <- cmultmodN(c=c, reg1=reg1, reg2=reg2,
               ancillas=ancillas, x=x, y=3, N=N)
z
z <- cmultmodN(c=c, reg1=reg1, reg2=reg2,
               ancillas=ancillas, x=z, y=3, N=N)
z
```

## Exponentiation modulo $N$

For the order finding algorithm, we have to implement the operation
$f_{a,N}(x) = a^x y \mod N$, where we set $y=1$ in the
following. All this is stored in a $n$ qubit register with $2^n > N$. 
Using the binary representation of $x$, we can write
\[
a^x\ =\ a^{2^0 x_0} \cdot a^{2^1 x_1} \cdot \ldots a^{2^{m-1}x_{m-1}}
\]
if $x$ is stored with $m$ bits.
So, if we start with the result register in state $|1\rangle$ we have
to multiply successivly $n$-times by $a^{2^i} \mod n$ depending on the
value of the qubit $|x_i\rangle$. This task can be achieved with the
controlled multiplier developed above.

Then we swap and multiply with the inverse pre-computed classically.

```{r}
basis <- c()
for(i in c(0:(2^13-1))) {
  basis[i + 1] <-
    paste0("|reg1=", i %/% (16*2^6) , ">|reg2=", i %/% (16*2^3) ,
           ">|reg3=", (i %/% 16) %% 2^3 ,
           ">|a=", (i %/% 8) %% 2, ">|c1=", (i %/% 4) %% 2,
           ">|c2=", (i%/%2) %% 2, ">|c=", i%%2, ">")
}
x <- X(9) * (X(1)*qstate(13, basis=basis))
c <- 1
ancillas <- c(2:4)
reg3 <- c(5:7)
reg2 <- c(8:10)
reg1 <- c(11:13)
N <- 5

```
