---
title: "Shor's Factoring Algorithm"
author: "Carsten Urbach"
output:
  rmarkdown::html_vignette
    
vignette: >
  %\VignetteIndexEntry{Shor's Factoring Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE}
library(knitr)
library(qsimulatR)
knitr::opts_chunk$set(fig.align='center',
                      comment='')
```

In order to break RSA cryptography one needs to be able to factorise a
large integer $n$, which is known to be the product of two prime
numbers $n=p q$. 

# Factoring Algorithm

Given an integer $n$, the factoring algorithm determines $p, q$ such
that $n=pq$. We assume $p,q\neq 1$.

1. Choose $m, 1\leq m \leq n$ uniformnly random with $m$ co-prime to $n$.
2. Find the order $r$ of $m$ modulo $n$.
3. If $r$ is even, compute $l=\mathrm{gcd}(m^{r/2}-1, n)$
4. If $l>1$ then $l$ is a factor of $n$. Otherwise, or if $r$ is odd
   start with 1 for another value of $m$.

## Greatest common divisor

Euclid described a classical algorithm for finding the greatest
common divisor (gcd) of two positive integers $m > n$.
It may be implemented recursively as follows:

```{r}
gcd <- function(m, n) {
  if(m < n) {
    return(gcd(m=n, n=m))
  }
  r <- m %% n
  if(r == 0) return(n)
  return(gcd(m=n, n=r))
}
```

## Order finding

Another ingredient is the order finding algorithm, which we are also
going to solve classically here, actually with the most naive algorithm

```{r}
findOrder <- function(x, n) {
  stopifnot(x < n && x > 0)
  tmp <- x %% n
  for(r in c(1:n)) {
    if(tmp == 1) return(r)
    tmp <- (tmp*x) %% n
  }
  if(tmp == 1) return(r)
  return(NA)
}
```

## Factoring

Shor's algorithms can be implemented as follows

```{r}
factoring <- function(n) {
  for(i in c(1:20)){
    ## generate random number
    m <- sample.int(n=n, size=1)
    cat("m=", m, "\n")
    ## Check, whether m, n are co-prime
    g <- gcd(n,m)
    if(g != 1 ) return(g)
    else {
      ## find the order of m modulo n
      r <- findOrder(x=m, n=n)
      cat("r=", r, "\n")
      if(!is.na(r)) {
        if((r %% 2) == 0) {
          l <- gcd(m^(r/2)-1, n)
          if(l > 1 && l < n) return(l)
        }
      }
    }
  }
  cat("could not find a factor!\n")
  return(NA)
}
```

And we can test whether it works

```{r, warning=FALSE}
set.seed(81)  ## for reproducibility
factoring(65)
factoring(91)
factoring(511)
```

Note that this computation is a bit tricky in `R` because of the
integer arithmetic with large integers. However, for our example here,
the code is sufficient.

